Supplementary Figures: The nature of decision noise in random exploration

### 2022-12-22 OLD/oldversions/Internal-External%20Noise/Internal-External Noise.tex
old version

	In this paper, we investigated whether random exploration is really random or whether it is driven deterministically by aspects of the stimulus we have previously ignored when measuring `decision noise'.  Using a version of the Horizon Task with repeated games, we found evidence that at least some of the noise in random exploration could be explained by such `deterministic noise.' In particular, we found that deterministic noise accounted for at least 27.05\% of the overall variability in people's behavior. %and increased with horizon - a hallmark of an exploratory process.

	So where does this leave randomness in random exploration? Well, the remaining 72.95 \% of the noise could be random or it could be deterministic because we can't control everything. In particular, while we controlled many aspects of the stimulus across repeated games (e.g. the outcomes and the order of the forced trials), we could not perfectly control {\it all} stimuli the participant received, which would vary, for example, based on exactly what they were looking at or whether they were scratching their nose. Thus conceptually, our estimate of deterministic noise is a lower bound. Conversely, our estimate of random noise is an upper bound as these `missing' sources of deterministic noise would be interpreted as random noise in our model. In addition, our estimation method also systematically underestimate deterministic noise (part of the deterministic noise will show up as random noise in model fitting, see Figure \ref{fig:pararecover2}), so methodologically our model also provides a lower bound of deterministic noise and an upper bound on random noise. Although there is still a considerable window for truly stochastic processes in the brain to be driving random exploration (up to 72.95\%), our results suggest that at least some of the apparent randomness in random exploration is not random at all. 
			%Taken at face value, the horizon-dependent increase in random noise is consistent with the idea that random exploration is driven by intrinsic variability in the brain.

	Randomness in exploratory behavior in random exploration can be driven by both the intrinsic variability generated in the brain (show up in behavior as random noise) and by deterministic neural processes induced by the stimuli (show up in behavior as deterministic noise).  The idea that random exploration is driven by truly stochastic processes in the brain is in line with work in the bird song literature in which variability during song learning has been tied to neural variability arising from specific areas of the brain \citep{songbird1, songbird2}. In a recent report from \cite{ebitz17}, the behavioral variability of monkeys in an `explore' state was also tied to random rather than deterministic sources of noise. In our study, most of the decision noise seen can still be random noise (up to 72.95\%), and this is consistent with findings of \cite{drugowitsch16}. In particular these authors show that randomness in behavior arises from imperfections in mental inference, that happen inside the brain, rather than in peripheral processes such as sensory processing and response selection. This suggests that a large proportion of noise in behavior is generated randomly and that this may arise from computational errors in computing the correct strategy. % In the context of the Horizon Task, such computational errors would likely be larger in the long horizon condition as the correct course of action in these cases is much harder to compute.
Whether such a noise-controlling area exists in the human brain is less well established, but one candidate theory \citep{aj2005} suggests that norepinephrine (NE) from the locus coeruleus may play a role in modulating random levels of noise. Indeed, changes in the NE system have been associated with changes behavioral variability in both humans and other animals in a variety of tasks \citep{eeKarpova14, eeKeung18}.  In addition there is some evidence that NE plays a direct role in random exploration \citep{eeWarren17}, although this finding is complicated by other work showing no effect of NE drugs on exploration \citep{jepma2012, nieuwenhuis05}.

	On the other hand, our result suggests that at least a quarter of the decision noise in random exploration arises from a deterministic process. This is in line with works of \citep{Bair1996} in which MT neurons were shown to have a reproducible temporal modulation in response to a fixed random motion stimuli. In other words, irrelevant features in the stimuli are represented in a reliable way in the brain that could drive downstream choices in a predictable way.  Of particular interest was the fact that deterministic noise in our study increased with horizon, which is a hallmark of an exploratory process and suggests that the modulation of deterministic processes may underlie random exploration.

@@

	Regardless of whether the remaining 72.95 \% is deterministic or random, the fact that the horizon change in the two noises are proportional to each other suggests a possible mechanism for random exploration. Specifically, a reduction in the strength with which reward drives the choice. In particular, in our analysis we excluded a scalar on $\Delta R$, if we include that then we get

	%REGARDLESS OF WHETHER THE REMAINING 75\% IS RANDOM OR NOT, THE FACT THAT THE HORIZON CHANGE IN THE TWO NOISES ARE PROPORTIONAL TO EACH OTHER SUGGESTS A POSSIBLE MECHANISM FOR RANDOM EXPLORATION.  SPECIFICALLY, A REDUCTION IN THE STRENGTH WITH WHICH REWARD DRIVES THE CHOICE.  IN PARTICULAR, IN OUR ANALYSIS WE EXCLUDED A SCALAR ON $\Delta R$.  IF WE INCLUDE THAT THEN WE GET
	\begin{equation}
	    \Delta Q = \beta \Delta R + A \Delta I + b + n_{det} + n_{ran}
	\end{equation}

>>>>

	In this paper, we investigated whether random exploration is really random or whether it is driven deterministically by aspects of the stimulus we have previously ignored when measuring `decision noise'.  Using a version of the Horizon Task with repeated games, we found evidence that at least some of the noise in random exploration could be explained by such `deterministic noise.' In particular, we found that deterministic noise accounted for at least 27\% of the overall variability in people's behavior. %and increased with horizon - a hallmark of an exploratory process.

	So where does this leave randomness in random exploration?  Is the remaining 73 \% of the noise that we cannot attribute to deterministic noise really random?  Or is this noise also deterministic, perhaps driven by factors we could not control in the lab?  In particular, while we controlled many aspects of the stimulus across repeated games (e.g. the outcomes and the order of the forced trials), we could not perfectly control {\it all} stimuli the participant received, which would vary, for example, based on exactly what they were looking at or whether they were fidgeting or scratching their nose CITE ANNE CHURCHLAND FIGETING (https://escholarship.org/content/qt1v23n1j7/qt1v23n1j7.pdf). Thus conceptually, our estimate of deterministic noise is a lower bound. Conversely, our estimate of random noise is an upper bound as these `missing' sources of deterministic noise would be interpreted as random noise in our model. Clearly, future work should try to control for these additional sources of deterministic noise as much as possible, for example by tracking people's pose, body movements and eye movements during the experiment.

	In addition to this conceptual limitation in measuring deterministic noise, parametere recovery simulations of our own experiment  suggest that our estimation method also systematically underestimate deterministic noise in this task (BY A FACTOR OF WHAT) (part of the deterministic noise will show up as random noise in model fitting, see Figure \ref{fig:pararecover2}).

	Thus, both conceptually and methodologically our model provides a lower bound of deterministic noise and an upper bound on random noise. Although there is still a considerable window for truly stochastic processes in the brain to be driving random exploration (up to 73\%), our results suggest that at least some of the apparent randomness in random exploration is not random at all. 
			%Taken at face value, the horizon-dependent increase in random noise is consistent with the idea that random exploration is driven by intrinsic variability in the brain. 
CUT FROM HERE
    If random exploration is random 
	Apparent randomness in random exploration can be driven by both the intrinsic variability generated in the brain (show up in behavior as random noise) and by deterministic neural processes induced by the stimuli (show up in behavior as deterministic noise).  The idea that random exploration is driven by truly stochastic processes in the brain is in line with work in the bird song literature in which variability during song learning has been tied to neural variability arising from specific areas of the brain \citep{songbird1, songbird2}. In a recent report from \cite{ebitz17}, the behavioral variability of monkeys in an `explore' state was also tied to random rather than deterministic sources of noise. In our study, most of the decision noise seen can still be random noise (up to 73\%), and this is consistent with findings of \cite{drugowitsch16}. In particular these authors show that randomness in behavior arises from imperfections in mental inference, that happen inside the brain, rather than in peripheral processes such as sensory processing and response selection. This suggests that a large proportion of noise in behavior is generated randomly and that this may arise from computational errors in computing the correct strategy. % In the context of the Horizon Task, such computational errors would likely be larger in the long horizon condition as the correct course of action in these cases is much harder to compute.
Whether such a noise-controlling area exists in the human brain is less well established, but one candidate theory \citep{aj2005} suggests that norepinephrine (NE) from the locus coeruleus may play a role in modulating random levels of noise. Indeed, changes in the NE system have been associated with changes behavioral variability in both humans and other animals in a variety of tasks \citep{eeKarpova14, eeKeung18}.  In addition there is some evidence that NE plays a direct role in random exploration \citep{eeWarren17}, although this finding is complicated by other work showing no effect of NE drugs on exploration \citep{jepma2012, nieuwenhuis05}.

	On the other hand, our result suggests that at least a quarter of the decision noise in random exploration arises from a deterministic process. This is in line with works of \citep{Bair1996} in which MT neurons were shown to have a reproducible temporal modulation in response to a fixed random motion stimuli. In other words, irrelevant features in the stimuli are represented in a reliable way in the brain that could drive downstream choices in a predictable way.  Of particular interest was the fact that deterministic noise in our study increased with horizon, which is a hallmark of an exploratory process and suggests that the modulation of deterministic processes may underlie random exploration.  
END CUT HERE

@@

	Regardless of whether the remaining 73 \% is deterministic or random, the fact that the horizon change in the two noises are proportional to each other suggests a possible mechanism for random exploration. Specifically, a reduction in the strength with which reward drives the choice.

	To see this, we need to return to our equation for the computation of value
		\begin{equation}
	    \Delta Q = \Delta R + A \Delta I + b + n_{det} + n_{ran}
	\end{equation}
	In this version of the equation, the difference in value is a weighted of sum of reward, information, bias, deterministic noise,  and random noise.  Crucially, we assumed that the weighting given to the difference in reward is 1 --- that is the multiplier on $\Delta R$ is 1, unlike the variable $A$ which modulates the effect of information.

	A more general form of this equation would also include a scalar on $\Delta R$

	%REGARDLESS OF WHETHER THE REMAINING 75\% IS RANDOM OR NOT, THE FACT THAT THE HORIZON CHANGE IN THE TWO NOISES ARE PROPORTIONAL TO EACH OTHER SUGGESTS A POSSIBLE MECHANISM FOR RANDOM EXPLORATION.  SPECIFICALLY, A REDUCTION IN THE STRENGTH WITH WHICH REWARD DRIVES THE CHOICE.  IN PARTICULAR, IN OUR ANALYSIS WE EXCLUDED A SCALAR ON $\Delta R$.  IF WE INCLUDE THAT THEN WE GET
	\begin{equation}
	    \Delta Q = \beta \Delta R + A \Delta I + b + n_{det} + n_{ran}
	\end{equation}
	In this equation, $\beta$ determines the weight given to the reward in the decision.  From a model fitting perspective, these two forms are EQUIVALENT. From a psychological perspective they are quite different.  And you can change the relative contribution of noise by changing $\beta$.

### 2022-12-22 OLD/oldversions/supplementary/Internal-External Noise.tex
old version

>>>>

	\begin{figure}[hp]
		\begin{center}
			\includegraphics[width=1\textwidth]{figures/precover.png}
			\caption{Parameter recovery over the subject-level means of information bonus, $A$, spatial bias, $b$, random noise variance, $\sigma_{ran}$, and deterministic noise variance, $\sigma_{det}$, for horizon 1 (left column) and horizon 6 (right column) games. 
			}
			\label{fig:pararecover1}
		\end{center}
	\end{figure}

	\begin{figure}[hp]
		\begin{center}
			\includegraphics[width=1\textwidth]{figures/RanDetNoise_pure_ran_det_simu_hyperprior.png}
			\caption{Parameter recovery over the subject-level means of information bonus, $A$, spatial bias, $b$, random noise variance, $\sigma_{ran}$, and deterministic noise variance, $\sigma_{det}$, for horizon 1 (left column) and horizon 6 (right column) games. 
			}
			\label{fig:pararecover1}
		\end{center}
	\end{figure}
